Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-08-31 doc/papers/cgo2014/paper.tex
CGO: Update section on existing work

%%
%%
\section{Introduction}
% On aura pas le temps pour le papier je pense mais faire une comparaison avec du
% C pure pour voir ce que les compilo de C arrive a faire.

The Python language~\cite{rossum97} now provides a full toolkit to carry out
scientific experiments~\cite{scipy}, from core scientific routines with the
Numpy package, to scientific packages with the Scipy package, plotting
facilities with the Matplotlib package, enhanced terminal and notebooks with
IPython. As a consequence, there has been a move from historical languages like
Fortran to Python, as showcased by the success of the Scipy conference.

@@

The success of Python and the development of its large ecosystem triggered
several attempts to improve its execution speed. The most significant basic
block for high performance numerical simulations is certainly Numpy, a
numerical library to manipulate dense arrays. Numerous routines are exposed in
Python and implemented in C, offering close-to-native performance in some cases.
% FIXME: Numpy is part of this section but somehow covered earlier???
%
However, this library-like approach has inherent limitations that were exposed
in the previous section. To address these limitations, several compiler-based
approaches were proposed.

\subsection{Numpy}
\cite{oliphant2007,numpyarray2011}

% Not sure what to add here because of the previous section.

\subsection{Cython}
\cite{cython2010}

%
Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing information, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variables to improve performance. Without type annotations, the
performance improvement is not terrific, but given enough type annotation,
Cython can generates code that runs as fast as its C equivalent, while
maintaining a syntax close to Python.

\subsection{Numba and Parakeet}

The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}~\cite{numba}
compiler uses additional type information to generate sequential LLVM bytecode.
Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
Python to LLVM bytecode translation, but limits its scope to the numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between numpy functions and a set
of parallel primitives including \texttt{map}s \texttt{scan}s and
\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
not suffer from backward-incompatibility issues. When meeting an unsupported
construct, numba falls back to Python C-API calls, while parakeet raises an
exception.

\subsection{Copperhead}
\cite{copperhead2011}

@@

\subsection{PyPy}
\cite{pypy2009}

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
enhance Python performance. However Shed~Skin does not provide support for
fine-grained parallelism beyond what the standard library proposes. PyPy is
heading toward STM for parallelism support.

>>>>

%%
%%
\section{Introduction}
% On aura pas le temps pour le papier je pense mais faire une comparaison avec
% du C pure pour voir ce que les compilo de C arrive a faire.

The Python language~\cite{rossum97} now provides a full toolkit to carry out
scientific experiments~\cite{scipy}, from core scientific routines with the
Numpy package\cite{oliphant2007,numpyarray2011}, to scientific packages with
the Scipy package, plotting facilities with the Matplotlib package, enhanced
terminal and notebooks with IPython. As a consequence, there has been a move
from historical languages like Fortran to Python, as showcased by the success
of the Scipy conference.

@@

The success of Python and the development of its large ecosystem triggered
several attempts to improve its execution speed. The most significant basic
block for high performance numerical simulations is certainly Numpy, a
numerical library to manipulate dense arrays. However, this library-like
approach has inherent limitations that were exposed in the previous section.
To address these limitations, several compiler-based approaches were proposed.

\subsection{Numpy Limitations}

Numpy programs typically makes use of high-level programming, mostly using
array broadcasting techniques as shown listing~\ref{lst:pairwise}. This level
of programming can be very efficiemt with Numpy when compared to regular Python
code. However the main limitation is that the Numpy library approach comes with
numerous temporary arrays allocated implicitely.

\begin{lstlisting}[language=python, caption={\textit{pairwise} function that
  exhibits array broadcasting.}, label={lst:pairwise}, breaklines=true]]
def pairwise_numpy(X):
    return np.sqrt(((X[:, None, :] - X) ** 2).sum(-1))
\end{lstlisting}

A solution to avoid temporary allocation can be to revert to a programming style
using explicit raw loops as shown listing~/ref{lst:pairwise-raw}. Unfortunately
this code runs almost 100 times slower due to the Python interpreter overhead.

\begin{lstlisting}[language=python, caption={\textit{pairwise} function
  using raw loops.}, label={lst:pairwise-raw}]
def pairwise_python(X):
    M = X.shape[0]
    N = X.shape[1]
    D = np.empty((M, M), dtype=np.float)
    for i in range(M):
        for j in range(M):
            d = 0.0
            for k in range(N):
                tmp = X[i, k] - X[j, k]
                d += tmp * tmp
            D[i, j] = np.sqrt(d)
    return D
\end{lstlisting}

\begin{lstlisting}[language=python]

\end{lstlisting}

\subsection{Cython}
%
Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing information, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variables to improve performance. Without type annotations, the
performance improvement is not terrific, but given enough type annotation,
Cython can generates code that runs as fast as its C equivalent.

The drawback is that the programmer has to avoid \textit{Pythonic}-like
high-level constructs and instead expresses his algorithm using raw loops.
The \textit{pairwise} function rewritten for Cython is shown 
listing~\ref{lst:pairwise-cython}.

\begin{lstlisting}[language=python, caption={\textit{pairwise} function 
  rewritten for Cython.}, label={lst:pairwise-cython}]
%%cython
import numpy as np
cimport cython
from libc.math cimport sqrt

@cython.boundscheck(False)
@cython.wraparound(False)
def pairwise_cython(double[:, ::1] X):
 cdef int M = X.shape[0]
 cdef int N = X.shape[1]
 cdef double tmp, d
 cdef double[:, ::1] D
 D = np.empty((M, M), dtype=np.float64)
 for i in range(M):
  for j in range(M):
   d = 0.0
    for k in range(N):
     tmp = X[i, k] - X[j, k]
     d += tmp * tmp
    D[i, j] = sqrt(d)
 return np.asarray(D)
\end{lstlisting}

\subsection{Numba}

The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}~\cite{numba}
compiler uses additional type information to generate sequential LLVM bytecode.
Numba uses Just-In-Time(JIT) compilation and do not suffer from
backward-incompatibility issues. When meeting an unsupported construct, numba
falls back to Python C-API calls.

Unfortunately, at that time Numba is providing its best speedup on code with 
raw loops, which is not encouraging programmers to express their code using 
the higher-level construct available with Numpy.

\subsection{Parakeet}

Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is
Python to LLVM bytecode translation, but limits its scope to the Numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between Numpy functions and a set
of parallel primitives including \texttt{map}s \texttt{scan}s and
\texttt{reduce}s.

\subsection{Copperhead}

@@

\subsection{PyPy}

PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
enhance Python performance. However Shed~Skin does not provide support for
fine-grained parallelism beyond what the standard library proposes. PyPy is
heading toward STM for parallelism support.
